---
title: "EDA"
author: "Timur Guler"
date: "8/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(gridExtra)
library(corrplot)
library(ROCR)
```


Report:
EXECUTIVE SUMMARY 
  Our team began this project by trying to answer a single question. Is it possible to use the physical and chemical properties of red and white wine to predict whether or not the wine is high quality? We took the perspective of a team of data scientists who was hired to advise a boutique wine seller. This seller is not interested in purchasing a high volume of wine, but is very interested in ensuring that they can get the highest ratio of good wines to bad wines in every one of their purchases. This is a difficult task considering there’s far more bad wine in the world than good wine. 
  Going into the project the team was already skeptical that we would be able to predict both red and white wine quality in one convenient model. Our limited knowledge of wine wasn’t enough for us to make this important decision right off the bat, so we needed to dig deeper. One crucial detail that we noticed is that although red and white wine share the same characteristics, during the initial exploration of the data we saw that red and white tend to have drastically different values in several of these categories. Some of these characteristics include residual sugar, fixed acidity, and chlorides. Additionally, some characteristics, like fixed acidity, affected quality differently for red and white wines. These factors indicated to our team that red and white wine deserved their own models in order to best meet our client’s needs. 
  The model building process was an iterative one where we tried to compare several different combinations of the wine characteristics in order to most effectively predict high versus low quality wine. A combination of exploratory data analysis and model selection criteria methods led us to select an initial 7-predictor model for both red and white wines. These initial models were validated through diagnostic procedures to ensure the logistic regression assumptions were satisfied. The significance of these models was then tested against other potential models, which included the intercept-only and full-predictor models. Several candidate models were compared by generating ROC curves, AUC scores and confusion matrices using test sets to determine which model best served our research question for both red and white wines. The best model for predicting high-quality red wine that we could create, while still keeping simplicity in mind, utilized seven different characteristics as predictor variables. Some of the more important predictors include alcohol content, volatile acidity, and sulphates. When we use this model to predict high-quality red wine, it correctly classifies a wine as high-quality 71% of the time. The best model for predicting high-quality white wine ended up using five predictors to classify the wine. Some of the most important predictors for this model had some overlap with the red wine model, like alcohol and volatile acidity, but then began to differ as we added in factors like residual sugar and pH. When this model classifies a white wine as high-quality, the model is correct slightly more than 60% of the time. While this is not as strong of a result as our red wine model, this is still extremely useful to our client. 
  While our models don’t predict high quality wines perfectly, it is an incredible supplement to human judgement alone. We recommend that our client use our model in conjunction with their years of experience with wine. We believe that marrying these two sources of information will lead to a much higher proportion of high-quality wines in stock, which in turn will lead to happy, loyal customers and less money wasted on low-quality wines. 
  In the future, our team would love to add more characteristics to our wine datasets that aren’t limited to the wine’s physical and chemical properties. These new predictive variables could include what region the wine comes from, what season it was produced in, and age of the wine. The model that we built isn’t only useful to our current client. This information could be informative for a wine producer to see if there are ways they can tweak the wine’s characteristics in order to engineer a higher-quality wine. 

**EDA**
Our goals: Our team took on the business challenge of advising a boutique wine store who wishes to identify high-quality red and white wines based on their physical and chemical characteristics. As a selective buyer, the agent wishes to only select the upper echelon of wines, and to minimize the number of low quality wines accidentally purchased. In addition to selecting high quality wines, the purchasing agent also wishes to learn which factors determine wine quality, and if and how these differ between white and red wines.

Our team used two data sets to answer these questions. One contained physical and chemical characteristics of red wine (1599 observations), and the other contained physical and chemical characteristics of white wine (4898 observations). This wine type imbalance, while not large enough to prevent meaningful analysis, was critical to keep in mind throughout our process.

Each data set contained 13 variables - an index column, a discrete 1-10 measure of quality, and the following physical/chemical characteristics of the wine:

* fixed acidity
* volatile acidity
* citric acidity
* residual sugar
* chlorides
* free sulfur dioxide
* total sulfur dioxide
* density
* pH
* sulphates
* alcohol

We removed the index column, as well as free sulfur dioxide, due to its high correlation with total sulfur dioxide in both red and white wine. Next, we performed a test-train split for both red and white wines with a fixed seed of 1, using 80% of each group for training data and the remaining 20% as testing data. Further discussions of "the data" during the EDA, model building, and model assumption validation stages will refer to the training data.

One fundamental issue in our analysis was the determination of a cutoff point for "high quality" wine. Based on the distribution for both red and white wines, we picked a cutoff quality rating of 7. This encompassed 13.5% of red wines and 21.4% of white wines. The same cutoff was used for both wine types for several reasons:

1. These proportions represent a level of distinction without making rarity a statistical concern
2. No determination had been made at this stage as to whether separate models would be used for white and red wine
3. While there was a statistically significant difference in mean quality levels (p=2.2e-16), the actual difference was not practically meaningful (5.877909 for whites vs.  5.636023 for reds).

We also added the variable "color" to identify the wines, and created a combined training data set in addition to the separate red and white training sets.

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
white_wine <- read.csv('wineQualityWhites.csv')
red_wine <- read.csv('wineQualityReds.csv')

white_wine <- dplyr::select(white_wine, -c(free.sulfur.dioxide, X))
red_wine <- dplyr::select(red_wine, -c(free.sulfur.dioxide, X))

red_wine <- red_wine %>%
  mutate(color = 0) #red is 0
white_wine <- white_wine %>%
  mutate(color = 1) #white is 1

#add quality_cat
red_wine <- red_wine %>%
  mutate(quality_cat = ifelse(quality>=7, 1, 0)) #high quality 1, low quality 0
white_wine <- white_wine %>%
  mutate(quality_cat = ifelse(quality>=7, 1, 0)) #high quality 1, low quality 0

#change color to factor
red_wine$color<-factor(red_wine$color)
levels(red_wine$color) <- c('red') 
white_wine$color<-factor(white_wine$color)
levels(white_wine$color) <- c('white') 

#change quality_cat to factor 
red_wine$quality_cat<-factor(red_wine$quality_cat)
levels(red_wine$quality_cat) <- c("low","high") 
white_wine$quality_cat<-factor(white_wine$quality_cat)
levels(white_wine$quality_cat) <- c("low","high") 

wine_combined <- rbind(white_wine, red_wine) #combine datasets 

#split into train and test data
set.seed(1) #set seed 
#white
sample_white<-sample.int(nrow(white_wine), floor(.80*nrow(white_wine)), replace = F)
train_white<-white_wine[sample_white, ] #training data 
test_white<-white_wine[-sample_white, ] #testing data

#red
sample_red<-sample.int(nrow(red_wine), floor(.80*nrow(red_wine)), replace = F)
train_red<-red_wine[sample_red, ] #training data 
test_red<-red_wine[-sample_red, ] 

#combined
train_combined <- rbind(train_white, train_red)
test_combined <- rbind(test_white, test_red) 

```

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plt1 <- ggplot(data = red_wine, aes(y=quality)) +
  geom_histogram(fill='indianred1', color='indianred4', alpha=0.75)
plt2 <- ggplot(data = white_wine, aes(y=quality)) +
  geom_histogram(fill='lightgoldenrod1', color='lightgoldenrod4', alpha=0.75)
grid.arrange(plt1, plt2, ncol=2)

```

Our team had several goals during the EDA process. Our first goal was to understand the relationships between variables for both red and white wines. We accomplished this by examining correlations and scatterplots.

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
par(oma=c(0,0,2,0))
corrplot(cor(train_red[, c(-12, -13)]), method="color", title = "Red Wine Correlations", outer=T)
corrplot(cor(train_white[, c(-12, -13)]), method="color", title="White Wine Correlations", outer=T)
```
This shows us that correlations between variables differ somewhat between white and red wines, and that each type of wine has some strong relationships between variables, which we will need to consider when looking for multicollinearity (e.g. fixed acidity with pH, density, and citric acid for red wines, density with residual sugar and alcohol in white wines). Looking at these relationships in more detail with scatterplots, we see that they are fairly tight, and somewhat linear.

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
pairs(train_red[, c("fixed.acidity", "pH", "density", "citric.acid", "quality")], lower.panel=NULL, main="Red Wine Scatterplots")
pairs(train_white[, c("density", "residual.sugar", "alcohol", "quality")], lower.panel=NULL, main="White Wine Scatterplots")
```

In addition to a general understanding of the variables' distributions and their relationships with each other, we were specifically interested in:

* the relationships between the physical/chemical characteristics of wine and quality 
* whether it would be more effective to build a combined model or separate models for red and white

First, we looked at scatterplots of quality vs. physical characteristic by color using best-fit regression lines, as well as boxplots of quality category vs physical characteristic by color. This led to three key takeaways:

1. While several characteristics seem to have relationships with quality, a visual analysis show low "tightness of fit" for each potential predictor, meaning that it will likely take the combination of several weak predictors to yield a good prediction model

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(data=train_combined, aes(x=alcohol, y=quality, color=color))+
  geom_point(alpha=0.3)+
  geom_smooth(method='lm')+
  labs(title="Quality by Alcohol and Color")
```


2. Some physical characteristics (notably fixed acidity, citric acidity, chlorides, pH, sulphates) have differing effects on quality in red vs white wines

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(data=train_combined, aes(x=fixed.acidity, y=quality, color=color))+
  geom_point(alpha=0.3)+
  geom_smooth(method='lm')+
  labs(title="Quality by Fixed Acidity and Color")
```


3. The range for some physical characteristics (notably residual sugar) is markedly different between red and white wines, leading to potential extrapolation issues if a combined model is used.

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(data=train_combined, aes(x=quality_cat, y=residual.sugar, color=color))+
  geom_boxplot()+
  facet_wrap(~color)+
  labs(title="Differences in the Effect of Residual Sugar on Quality, by Color")
```

Based on takeaways (2) and (3), as well as the difference in sizes between the red and white data sets, our team determined that it would be best to use two separate models for predicting high vs. low quality - one for red wines, and one for white wines.


red wine model:

- various measures point to 7 predictor model
- fit model and review assumptions (influential points, linearity assumption, multicollinearity)

- p value of 7 predictor vs intercept only - useful

- 7 more useful than full model

----------------test data--------------
- look at roc, auc

- confusion matrix - show multiple thresholds, end up achieving desired business goals

white wine model:

repeat process

------------discussion------------------

relate back to research question

